---
title:  fGSEA
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_document:
    code_folding: hide
    theme: space
    toc: yes
    toc_depth: 3
    toc_float: no
  BiocStyle::html_document2:
    code_folding: hide
    toc: yes
    toc_float: yes
  knitrBootstrap::bootstrap_document:
    highlight.chooser: yes
    theme.chooser: yes
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, bootstrap.show.code = FALSE, results='hide', bootstrap.show.message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE}
knitr::opts_chunk$set(bootstrap.show.code = FALSE, message=FALSE, warning=FALSE)
suppressMessages(library(DESeq2))
suppressMessages(library(data.table))
suppressMessages(library(magrittr))
suppressMessages(library(QoRTs))
suppressMessages(library(reshape2))
suppressMessages(library(RColorBrewer))
suppressMessages(library(ggplot2))
suppressMessages(library(tidyr))
suppressMessages(library(plyr))
suppressMessages(library(kableExtra))
suppressMessages(library(gtools))
suppressMessages(library(plotly))
suppressMessages(library(gridExtra))
suppressMessages(library(openxlsx))
suppressMessages(library(pathview))
suppressMessages(library(edgeR))
suppressMessages(library(limma))
suppressMessages(library(openxlsx))
suppressMessages(library(pheatmap))
suppressMessages(library(gplots))
suppressMessages(library(treemap))
suppressMessages(library(scales))
suppressMessages(library(Hmisc))
suppressMessages(library(knitr))
suppressMessages(library(ggrepel))
suppressMessages(library(fgsea))
suppressMessages(library(clusterProfiler))
tol12qualitative=c("#9e9ac8", "#6a51a3", "#fdbe85", "#fd8d3c", "#d94701", "#74c476", "#41ab5d", "#238b45", "#005a32", "#bdd7e7", "#6baed6", "#2171b5")
palette <- c("#d94701", "#6a51a3", "#2171b5", "#238b45")
```



```{r readInData, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
counts <- read.table(file = "gene.counts.txt", header = TRUE, check.names=FALSE, row.names=1)
decoderFile <- "decoder.txt"
decoder.data <- fread(decoderFile) %>% as.data.frame()
decoder.data$group <- factor(decoder.data$group, levels=c("TCR.OTI", "TCR.OTI.SMARTA"))
decoder.data <- decoder.data[decoder.data$sample.ID %in% colnames(counts),]
counts <- counts[,c(decoder.data$sample.ID)]
if(!identical(decoder.data$sample.ID, colnames(counts))) stop()
```

## Samples

```{r printSampleTable, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
kable(decoder.data, row.names=FALSE,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Differential expression analysis


```{r runDESeq, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
coldata <- data.frame(decoder.data, row.names = colnames(counts), stringsAsFactors=F)
coldata$group <- factor(decoder.data$group)
dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts,
                                      colData = coldata,
                                      design = ~group)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)
vsd <- DESeq2::vst(dds, blind=TRUE)

alphaTR = 0.05

OT1.SMARTA_vs_OT1 <- results(dds, contrast=c("group","TCR.OTI.SMARTA", "TCR.OTI"), alpha=alphaTR)
```



The following genes were detected as differentially expressed (adjusted p < `r alphaTR`):


```{r printDEGtable, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE, eval=TRUE}
de.all = list(OT1.SMARTA_vs_OT1=OT1.SMARTA_vs_OT1)
de.all.sig <- lapply(de.all, subset, padj < alphaTR)

deg_table =  as.data.frame(sapply(de.all.sig, nrow))
colnames(deg_table) <- paste0("padj<",alphaTR)

kable(deg_table, row.names=T)  %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```


# GSEA w.r.t. 

gene sets from table s6  https://www.science.org/doi/10.1126/science.abb9847 


```{r subchunkify, echo=FALSE}
#' https://github.com/andschar/dbreport/blob/546410a9aefc597bf78d01bb9dae0de4448439f6/R/utils.R
#' Create sub-chunks for plots
#'
#' taken from: https://stackoverflow.com/questions/15365829/dynamic-height-and-width-for-knitr-plots
#'
#' @param pl a plot object
#' @param fig.height figure height
#' @param fig.width figure width
#' @param chunk_name name of the chunk
#'
#' @author Andreas Scharmueller \email{andschar@@protonmail.com}
#'
subchunkify = function(pl,
                       fig.height = 7,
                       fig.width = 5,
                       chunk_name = 'plot') {
  pl_deparsed = paste0(deparse(function() {
    pl
  }), collapse = '')
  
  sub_chunk = paste0(
    "```{r ",
    chunk_name,
    ", fig.height=",
    fig.height,
    ", fig.width=",
    fig.width,
    ", dpi=72",
    ", echo=FALSE, message=FALSE, warning=FALSE}",
    "\n(",
    pl_deparsed,
    ")()",
    "\n```"
  )
  
  cat(knitr::knit(
    text = knitr::knit_expand(text = sub_chunk),
    quiet = TRUE
  ))
}
```

```{r read_in, eval=T,  message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE, eval=T}
library(stringr)
gene.sets =read.xlsx("abb9847_table_s6.xlsx")
lst <- as.list(gene.sets)

lst <- lapply(lst, function(x) {
  x <- str_to_title(tolower(x))
  x[!is.na(x)]
})

lst = lst[c("CD39-CD69-","CD39+CD69+")]
```

```{r fgsea_custom, fig.width=7, fig.height=12 , fig.align="center", results='asis'}
library(data.table)
library(magrittr)
library(ggplot2)

set.seed(149)
res <-  OT1.SMARTA_vs_OT1[!is.na(OT1.SMARTA_vs_OT1$padj),] %>% as.data.frame %>% as.data.table(., keep.rownames=TRUE)
setnames(res, "rn", "gene_symbol")
setorder(res, -stat)
geneList <- res$stat
names(geneList) <- res$gene_symbol

gene.sets <- lst

set.seed(149)
fgseaRes <- fgseaMultilevel(gene.sets, geneList, minSize=15, maxSize=500)
topPathways <- fgseaRes[padj < 0.05][order(NES), pathway]
fgseaResSig <- subset(fgseaRes, padj < 0.05)
mainPathways <- fgseaResSig[pathway %in% fgseaResSig$pathway][order(-NES), pathway]
fgseaResMain <- fgseaRes[match(mainPathways, pathway)]

lapply(fgseaResSig$pathway, function(y){
  cat("\n\n###",y, "\n\n")
  
  subchunkify(  plot(plotEnrichment(gene.sets[[y]],geneList) + labs(title=gsub("_", " ", y), subtitle=paste0("", "; NES ",round(fgseaResSig[which(fgseaResSig$pathway == y)]$NES, 2),"; ",paste0("adj. p < ",round(fgseaResSig[which(fgseaResSig$pathway == y)]$padj, 100)) ))), fig.width=4.5, fig.height=3, chunk_name=paste0(y,"gsea"))
  
  breaksList = seq(-2, 2, by = .1)
  selected_genes <-  fgseaResSig[fgseaResSig$pathway == y,]$leadingEdge[[1]]
  selected_genes_lbls <- selected_genes
  pheatmap(assay(vsd)[selected_genes,], scale="row",  show_rownames = TRUE,cluster_rows = TRUE, cluster_cols = F, gaps_col = c(4), main=y, fontsize_row = 12, col =  bluered(length(breaksList)), cellwidth=20, cellheight=10, breaks = breaksList, border_color = NA)
  plot.new() ;
}) %>%  invisible()
```




***
# Session Info
```{r session, message=FALSE, warning=FALSE, cache=TRUE,comment="",echo=FALSE, fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```

# References