---
title: Heatmap 
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output:
  html_document:
    code_folding: hide
    theme: space
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, bootstrap.show.code = FALSE, results='hide', bootstrap.show.message=FALSE, warning=FALSE, cache=TRUE}
suppressMessages(library(DESeq2))
suppressMessages(library(DiffBind))
suppressMessages(library(tidyverse))
suppressMessages(library(data.table))
suppressMessages(library(kableExtra))
suppressMessages(library(ggrepel))
suppressMessages(library(pheatmap))
suppressMessages(library(gplots))
suppressMessages(library(TxDb.Mmusculus.UCSC.mm10.knownGene))
suppressMessages(library(openxlsx))
suppressMessages(library(profileplyr))
suppressMessages(library(ComplexHeatmap))
suppressMessages(library(RColorBrewer))
set.seed(0813185)
```


```{r readInData, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
DB <- readRDS("2022_01.DB_gabriel.Rds")
db <- dba.count(DB, peaks=NULL, score=DBA_SCORE_READS)
consensus_peaks <- dba.peakset(db, bRetrieve=TRUE)
counts = as.data.frame(mcols(consensus_peaks))
row.names(counts) = paste0(as.data.frame(consensus_peaks)$seqnames, ":",as.data.frame(consensus_peaks)$start,"-",as.data.frame(consensus_peaks)$end)
decoderFile <- "atac_decoder.txt"
decoder.data <- fread(decoderFile) %>% as.data.frame()
decoder.data$group <- factor(decoder.data$group, levels=c("TCR.OTI", "TCR.OTI.SMARTA"))
decoder.data <- decoder.data[decoder.data$sample.ID %in% colnames(counts),]
counts <- counts[,c(decoder.data$sample.ID)]
if(!identical(decoder.data$sample.ID, colnames(counts))) stop()
```

## Samples

```{r printSampleTable, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
kable(decoder.data, row.names=FALSE,  padding = 0, longtable=TRUE) %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Differential accessibility analysis


```{r runDESeq, message=FALSE, warning=FALSE, cache=TRUE, context="data"}
coldata <- data.frame(decoder.data, row.names = colnames(counts), stringsAsFactors=F)
coldata$group <- factor(decoder.data$group)
dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts,
                                      colData = coldata,
                                      design = ~group)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds <- DESeq(dds)
vsd <- DESeq2::vst(dds, blind=TRUE)

alphaTR = 0.05

OT1.SMARTA_vs_OT1 <- results(dds, contrast=c("group","TCR.OTI.SMARTA", "TCR.OTI"), alpha=alphaTR)
```

The following genes were detected as differentially expressed (adjusted p < `r alphaTR`):


```{r printDEGtable, message=FALSE, warning=FALSE, cache=TRUE, echo=FALSE, eval=TRUE}
de.all = list(OT1.SMARTA_vs_OT1=OT1.SMARTA_vs_OT1)
de.all.sig <- lapply(de.all, subset, padj < alphaTR)

deg_table =  as.data.frame(sapply(de.all.sig, nrow))
colnames(deg_table) <- paste0("padj<",alphaTR)

kable(deg_table, row.names=T)  %>%  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width=F)
```



# Heatmap of DA peaks (kmeans = 2)

VST normalized expression of DA peaks.

```{r heatmaps, message=FALSE, warning=FALSE, cache=TRUE, fig.width=5, fig.height=5, context="data", fig.align='center'}

data = OT1.SMARTA_vs_OT1 %>% subset(., padj < alphaTR) %>% as.data.frame()

breaksList = seq(-1.5, 1.5, length = 100)

selected_features <- row.names(data)

log2.count.matrix = assay(vsd)[selected_features,]

#Obtain the z-score per peak region across all the samples
log2.count.matrix <- apply(log2.count.matrix, MARGIN = 1, FUN = function(X) (X - mean(X))/sd(X))
log2.count.matrix <- t(log2.count.matrix)

k=2
km <- kmeans(log2.count.matrix,centers=k, iter.max=500) 
m.kmeans<- cbind(log2.count.matrix, km$cluster) # combine the cluster with the matrix
o<- order(m.kmeans[,ncol(m.kmeans)]) # order the last column
colnames(m.kmeans)[ncol(m.kmeans)] <- "k"
m.kmeans = m.kmeans[o,]
annotation_row = data.frame(row.names=row.names(m.kmeans), km=as.factor(m.kmeans[,ncol(m.kmeans)])) 
ordered_mat = m.kmeans[,1:ncol(m.kmeans)-1]

pheatmap::pheatmap(ordered_mat, scale="none", show_rownames=F, fontsize = 10, cellwidth =25,  main = "DAPs", col=viridis::viridis(50), silent = F , cluster_rows=F, cluster_cols = F, annotation_row=annotation_row)

```

```{r saveHeatmapData, eval=T,  warning=FALSE, cache=TRUE, echo=FALSE,  fig.width=16, fig.height=10}
excel_df = ordered_mat
if(!identical(row.names(annotation_row), row.names(excel_df))) stop()

OT1.SMARTA_vs_OT1_sig = OT1.SMARTA_vs_OT1 %>% subset(., padj < alphaTR) %>% as.data.frame()

excel_df= merge(excel_df, OT1.SMARTA_vs_OT1_sig, by="row.names")
row.names(excel_df) = excel_df$Row.names
 excel_df$Row.names = NULL
excel_df= merge(excel_df, annotation_row, by="row.names")
colnames(excel_df)[1] = "peak"

spl <- strsplit(gsub("-|:", ".", excel_df$peak), ".", fixed = TRUE)
excel_df$chr = sapply(spl, "[", 1)
excel_df$start =sapply(spl, "[", 2)
excel_df$end = gsub("\\_.*","",sapply(spl, "[", 3))
excel_df_gr = GRanges(excel_df)
excel_df_gr_anot = ChIPseeker::annotatePeak(excel_df_gr,tssRegion = c(-1000,1000),TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene,annoDb = "org.Mm.eg.db", verbose=F)
excel_df_gr_anot_df = as.data.frame(excel_df_gr_anot)
row.names(excel_df_gr_anot_df) = excel_df_gr_anot_df$peak
excel_df_gr_anot_df = excel_df_gr_anot_df[row.names(ordered_mat), ]
if(!identical(row.names(excel_df_gr_anot_df), row.names(ordered_mat))) stop()

write.xlsx(excel_df_gr_anot_df, file="hm_atac_ordered.xlsx")
```


```{r prep_bw_hm, eval=T,  warning=FALSE, cache=TRUE, echo=FALSE,  fig.width=16, fig.height=10}
consensus_peaks_df = consensus_peaks
mcols(consensus_peaks_df) = NULL
consensus_peaks_df = as.data.frame(consensus_peaks_df)
row.names(consensus_peaks_df) = paste0(as.data.frame(consensus_peaks_df)$seqnames, ":",as.data.frame(consensus_peaks_df)$start,"-",as.data.frame(consensus_peaks_df)$end)

if(!identical(row.names(annotation_row), row.names(ordered_mat))) stop()

peaks_of_interest = consensus_peaks_df[row.names(ordered_mat),]
peaks_of_interest = merge(peaks_of_interest, annotation_row, by="row.names")
row.names(peaks_of_interest)  = peaks_of_interest$Row.names
peaks_of_interest$Row.names = NULL
peaks_of_interest = peaks_of_interest[row.names(ordered_mat),]

write.table(data.frame(peaks_of_interest[,c(1:3)], ID=row.names(peaks_of_interest), peaks_of_interest[,c(6)]), "hm_atac_ordered.bed", sep="\t", quote=F, col.names=F, row.names = F)

all_profiles =  BamBigwig_to_chipProfile(c("OT1.bw", "OT1.SMARTA.bw"),testRanges ="hm_atac_ordered.bed",style = "point",format = "bigwig",distanceAround = 1000)

# table(row.names(assay(all_profiles, 1)) == row.names(assay(all_profiles, 2)))
all = cbind(assay(all_profiles, 1),assay(all_profiles, 2))
row.names(all) = rowData(all_profiles)$name

all = all[row.names(peaks_of_interest),]

table(row.names(peaks_of_interest) == row.names(all))
#  saveRDS(all, "bw_hm_prep.Rds")
```

## Bigwig heatmap

To improve visibility, bins with read counts greater than the 75th percentile + 1.5*IQR were capped at that value.


```{r bw_heatmap_union_bg_clsrow_dend_reorder_f, eval=T,  warning=FALSE, cache=TRUE, echo=FALSE,  fig.width=3, fig.height=7}

#all = readRDS("bw_hm_prep.Rds")

# To improve visibility, bins with read counts greater than the 75th percentile + 1.5*IQR were capped at that value.
# (quantile(all, c(0, .75))+IQR(all)*1.5)[2]

all[all >= as.numeric((quantile(all, c(0, .75))+IQR(all)*1.5)[2])] <- as.numeric((quantile(all, c(0, .75))+IQR(all)*1.5)[2])

#table(row.names(annotation_row) == row.names(all))

annot_col =data.frame(cond =rep(gsub(".bw", "", metadata(all_profiles)$names), each=100),row.names=colnames(all))
annot_col$cond = factor(annot_col$cond, levels=unique(rep(gsub(".bw", "", metadata(all_profiles)$names), each=100)))

cols <- inlmisc::GetColors(n = 2, scheme="light")
rgb2hex <- function(r,g,b) sprintf('#%s',paste(as.hexmode(c(r,g,b)),collapse = ''))

ha =  ComplexHeatmap::HeatmapAnnotation(foo =  ComplexHeatmap::anno_block(gp = gpar(fill = c(rgb2hex(204,236,230),rgb2hex(198,219,239))), labels = unique(rep(gsub(".bw", "", metadata(all_profiles)$names), each=100)),labels_gp = gpar(col = c("black", "black"))))

#pdf("test.pdf")
hm = ComplexHeatmap::Heatmap(all, border=F, cluster_rows = T, cluster_columns = F, col=rev(brewer.pal(n = 11, name ='RdBu')), show_row_names = F, heatmap_legend_param = list(direction = "horizontal", title=NULL,legend_width = unit(6, "cm")), column_split = annot_col$cond, top_annotation = ha, column_title = NULL, show_column_names =F, use_raster=F, show_row_dend = F, row_split = peaks_of_interest$km, cluster_row_slices=F, row_title=NULL, row_gap = unit(0.5, "mm"), row_dend_reorder=F)  +  ComplexHeatmap::Heatmap(peaks_of_interest$km, name = "km", width = unit(3, "mm"), col = structure(cols, names = 1:2), row_split = peaks_of_interest$km, row_gap = unit(0.5, "mm"), show_heatmap_legend = F, border=T) 
ComplexHeatmap::draw(hm, heatmap_legend_side="bottom")
#dev.off()

hm = ComplexHeatmap::Heatmap(all, border=F, cluster_rows = T, cluster_columns = F, col=rev(brewer.pal(n = 9, name ='RdBu')), show_row_names = F, heatmap_legend_param = list(direction = "horizontal", title=NULL,legend_width = unit(6, "cm")), column_split = annot_col$cond, top_annotation = ha, column_title = NULL, show_column_names =F, use_raster=F, show_row_dend = F, row_split = annotation_row$km, cluster_row_slices=F, row_title=NULL, row_gap = unit(0, "mm"), row_dend_reorder=F)
ComplexHeatmap::draw(hm, heatmap_legend_side="bottom")

```


# Session Info
```{r session, message=FALSE, warning=FALSE, cache=TRUE,comment="", fig.width=10, fig.height=5.5, context="data"}
sessionInfo()
```
